Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-05-25 Paper/paper.tex
tov text

% \red{
% BEgin this paragraph by pointing out that EOS is universal among all NSs. 
% Therefore, one universal $\lambdans$ vs $\mns$ curve. Therefore, by combining
% multiple BHNS events, expect to measure $\lambdans$ better than with any 
% single BHNS event.
% }
% % \prayush{
The EoS of matter at nuclear densities is believed to be universal among all
neutron stars. The TOV equations therefore furnish a unique dependence of
$\lambdans$ on $\mns$. This fact enables us to go further - combine
information from multiple observations of realistic NSBH systems and
perform a fully-Bayesian analysis of how our estimation of $\lambdans$ changes
as we accumulate detections. This is similar to an earlier
study~\cite{DelPozzo:13} aimed at binary neutron stars.
% % }
% % 
% \red{
% MOVE into later sections. Keep here only 1-2 informational sentences.
% }
% % 
We restrict ourselves to a population of NSs with masses clustered very tightly
around $1.35M_\odot$ (with a negligible variance), and negligible spins. We
sample different nuclear EoSs by sampling entire populations with different
values set for the tidal deformability of the NS.
% These values correspond to a hard EoS at the lower end, and a soft EoS at the upper.
Finally, we take source
locations to be uniformly distributed in spatial volume, and source
orientations to be uniform on a $2-$sphere. From an ensemble of such
populations, we learn the following.
% 
Our best estimate for $\lambdans$ improves much faster than our statistical
uncertainties. Within $10-20$ detections, the median measured value of
$\lambdans$ lies within $10\%$ of its actual value.
Measurement uncertainties for $\lambdans$, on the other hand, depend on
$\lambdans$ itself. We find that for hard equations of state (with 
$\lambdans\geq 1000$), $10-20$ observations are sufficient to constrain
$\lambdans$ within $\pm 50\%$. For softer equations of state, the same level
of certainty would require substantially more ($25-40$) observations.
% 
Further, if the astrophysical ``mass-gap'' is real, we find that $20-50\%$
additional observations would be required to attain the same measurement
accuracy as above. Putter tighter constraints on the $\lambdans$ of a
population would require $50+$ NSBH observations, in either scenario.
% 
All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}.

@@

% \prayush{
In rest of the paper, we restrict our parameter space to span 
mass-ratios $q:=\mbh/\mns\in[2,5]$, black hole spin (aligned with orbit)
$\chibh\in[-0.5, +0.75]$, and dimensionless neutron star tidal deformability 
$\lambdans:= G\left(\frac{c^2}{G \mns}\right)^5\lambda \in[500, 2000]$.
The mass of NS for signals (and not templates) is fixed to $1.35M_\odot$ 
throughout~\cite{stellarcollapsemass}, and is taken to be 
non-spinning~\cite{Miller:2014aaa}.
Each signal is rescaled to a range of SNR values $\rho\in[10, 70]$.
% }
% 
Finally, we note that our results are applicable for the design sensitivity
LIGO instruments, that are expected to come online in $2018-19$.

>>>>

% \red{
% BEgin this paragraph by pointing out that EOS is universal among all NSs. 
% Therefore, one universal $\lambdans$ vs $\mns$ curve. Therefore, by combining
% multiple BHNS events, expect to measure $\lambdans$ better than with any 
% single BHNS event.
% }
% % \prayush{
The EoS of matter at nuclear densities is believed to be universal among all
neutron stars. The Tolman-Oppenheimer-Volkoff equation would then predict that all
NSs lie on a universal $\lambdans$ versus $\mns$ curve. As the final part of
this paper, we combine information from multiple observations of realistic NSBH
systems and perform a fully-Bayesian analysis of how our estimation of
$\lambdans$ changes as we accumulate detections. This is similar to an earlier
study~\cite{DelPozzo:13} aimed at binary neutron stars.
% % }
% % 
% \red{
% MOVE into later sections. Keep here only 1-2 informational sentences.
% }
% % 
We restrict ourselves to a population of NSs with masses clustered very tightly
around $1.35M_\odot$ (with a negligible variance), and negligible spins. We
sample different nuclear EoSs by sampling entire populations fixing different
values for the NS tidal deformability.
% These values correspond to a hard EoS at the lower end, and a soft EoS at the upper.
For all populations, we take source locations to be uniformly distributed in
spatial volume, and source orientations to be uniform on a $2-$sphere. To
summarize, we find the following:
% 
(a) Our best estimate for $\lambdans$ improves much faster than our statistical
uncertainties. Within $10-20$ detections, the median measured value of
$\lambdans$ lies within $10\%$ of the actual value.
(b) Measurement uncertainties for $\lambdans$, on the other hand, depend on
$\lambdans$ itself. We find that for hard equations of state (with 
$\lambdans\geq 1000$), $10-20$ observations are sufficient to constrain
$\lambdans$ within $\pm 50\%$. For softer equations of state, the same level
of certainty would require substantially more ($25-40$) observations.
% 
(c) Further, if the astrophysical ``mass-gap'' is real, we find that $20-50\%$
additional observations would be required to attain the same measurement
accuracy as above. And, (d) putting tighter constraints on the $\lambdans$ of a
population would require $50+$ NSBH observations, in any scenario.
% 
All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}.

@@

% \prayush{
In rest of the paper, we restrict our parameter space to span 
mass-ratios $q:=\mbh/\mns\in[2,5]$, black hole spin (aligned with orbit)
$\chibh\in[-0.5, +0.75]$, and dimensionless neutron star tidal deformability 
$\lambdans:= G\left(\frac{c^2}{G \mns}\right)^5\lambda \in[500, 2000]$.
The mass of NS for signals (and not templates) is fixed to $1.35M_\odot$ 
throughout~\cite{stellarcollapsemass}, and is taken to be 
non-spinning~\cite{Miller:2014aaa}.
Each signal is rescaled to a range of SNR values $\rho\in[10, 70]$.
% }
% 
Finally, we note that our results are applicable for the design sensitivity
LIGO instruments, that are expected to come online in $2018-19$.
\prayush{INSERT TEXT ON LIMITATIONS}
